Revealing the movement dynamics of hippocampal replay using state space modeling

### 2020-09-30 Article.tex
Update on Overleaf.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{Verifying the model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For the second 60 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 60 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (shading). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time. The summed posterior across states also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. The lines correspond to the average probability of that dynamic for a given constant velocity sequence. We use a 0.8  threshold (dotted line) to classify each sequence based on the state or states which contribute maximally to the posterior. 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 180 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 60 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 60 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v5}
\caption{Demonstration of the state space model on hippocampal data using both sorted and clusterless spikes on the same SWR event. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure \ref{1}F. Shaded regions correspond to the speed classifications as in Figure \ref{1}G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as (A).  \textbf{(C)} Maximum of the 1D decoded position (the most probable position) in (A) projected back onto the 2D maze for the sorted decoding. Color indicates time. The animal's actual position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire recording session. L, R, and C correspond to the left, right and center reward well as in (A). \textbf{(D)} Maximum of the 1D decoded position (the most probable position) in (B) projected back into 2D using the clusterless decoding. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v5}
\caption{A-D. Four examples of replays that would not be well-characterized by using the standard decoder. Figure conventions are the same as in Figure \ref{2}. \textbf{(A)} A replay that starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that persistently represents the left well for the entirety of the SWR while the animal remains in the center arm. \textbf{(C)} A replay that begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each day that has some content that falls into one of the categories of dynamics (Stationary, Stationary-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each day that have some Stationary, Stationary-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each day that has some Continuous content.
}
\label{3}
\end{figure*}

@@

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{95 \% Highest Posterior Density of the Dynamics. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. Top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. 2nd panel: Posterior distribution of estimated linear position over time. Red triangle indicates animal's position. 3rd panel: Posterior distribution plot, where black designated the position bins that are included in the 95\% highest posterior density (HPD) region. 4th panel: Cumulative HPD size. \textbf{(E)} Average Cumulative 95\% HPD size for each dynamic category.}
\label{4}
\end{figure*}

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each SWR. Each row represents a possible dynamic and each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total number of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of stationary trajectories on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of average tetrode multiunit spike rates for each dynamic category within each SWR.
}
\label{5}
\end{figure*}

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}
\caption{The model is robust to change of the probability of persisting in the same dynamic. Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value---which governs the probability of remaining in that state. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in Figure 1D. The first plot (upper left) shows the case when all dynamics are equally likely. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another state.}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental1/Figure2_v3-supplemental1}
\caption{Decoding the same SWR in Figure \ref{2} with 2D position using sorted spikes and clusterless decoding, respectively. \textbf{(A)} The left plot shows the spikes from cells arranged by the linear position of the peak of place field as in Figure 2. The middle plot shows the probability of each dynamic over time from the 2D decode. Shaded regions indicate classification category as in Figure 1G and 2. The rightmost plot shows the most probable estimate of the latent position (MAP estimate) with color indicating time. The latent position posterior summed over time is shown in purple underneath. The light grey lines represent the position of the animal over the entire epoch and the red dot represents the animal's position. \textbf{(B)} Same as in A, but with clusterless decoding.}
\label{fig:Figure2-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental2/Figure2_v4-supplemental2}
\caption{A-C. Examples of continuous replays. Left panel uses the same conventions as Figure 2A and 2B. Right panel shows the 1D MAP estimate projected back to 2D as in Figure 2C. Color indicates time. Light grey lines indicate the animal's position over the entire epoch.}
\label{fig:Figure2-Figure supplement 2}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental1/Figure3_v4_supplemental1}
\caption{More examples of replays that would not be well-characterized by using the standard decoder. Conventions are the same as in Figure 3. \textbf{(A)} A continuous replay that starts in the left arm back toward the center well, jumps to the right arm and continues back toward the center well. \textbf{(B)} A continuous replay that travels down the right arm toward the center well, proceeds past the choice point toward the left well and then returns back to the choice point. \textbf{(C)} A replay that starts down the left arm toward the center arm and turns into a stationary trajectory near the choice point. \textbf{(D)} The first half is a fragmented replay and the second half is unclassified. \textbf{(E)} A stationary trajectory in the first half of the SWR on the right arm and then a stationary trajectory in the SWR at the left well. \textbf{(F)} A stationary trajectory that starts at the animal's position at the center well, then jumps to the left well and comes back toward the center well.
}
\label{fig:Figure3-Figure supplement 1}
\end{figure*}

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental2/Figure3_v2-supplemental2}
\caption{Comparison of speed classification on real vs. position shuffled data for two epochs from different animals. Line represents the percent of SWR events classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing and position occupancy.}
\label{fig:Figure3-Figure supplement 2}
\end{figure}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4-supplemental1/Figure4_v1-supplemental1}
\caption{Comparison of average spatial coverage of the 95\% highest posterior density on real vs. position shuffled data for two epochs from different animals. Line represents the average spatial position spanned by the highest posterior density on real data for all ripples. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing and position occupancy.}
\label{fig:Figure4-Figure supplement 1}
\end{figure*}

>>>>

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time. The summed posterior across states also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors). 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 180 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 60 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 60 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v5}
\caption{The model can decode hippocampal replay trajectories using either sorted and clusterless spikes from the same SWR event. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure \ref{1}F. Shaded regions correspond to the speed classifications as in Figure \ref{1}G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the the magenta dashed line. Underneath is the maximum of the 1D decoded position (the most probable position) projected back onto the 2D track for the sorted decoding. Color indicates time. The animal's actual position is denoted by the pink dot. Light grey lines show the animal's 2D position over the entire recording session. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as (A). Underneath is the maximum of the 1D decoded position (the most probable position) projected back into 2D using the clusterless decoding.
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v5}
\caption{A-F. Six examples of SWRs that would not be well-characterized by using the standard decoder. Figure conventions are the same as in Figure \ref{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the choice point. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{3}
\end{figure*}

@@

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{Validation of the Dynamics using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. Top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. 2nd panel: Posterior distribution of estimated linear position over time. Red triangle indicates animal's position. 3rd panel: Posterior distribution plot, where black designated the position bins that are included in the 95\% highest posterior density (HPD) region. 4th panel: Cumulative HPD size. \textbf{(E)} Average Cumulative 95\% HPD size for each dynamic category.}
\label{4}
\end{figure*}

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Quantification of the SWRs by dynamic. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Box plot of the duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Box plot of the average distance between the latent position to the animal's position for each classification within the SWR. \textbf{(D)} Box plot of the average speed of the classification within the SWR (excluding classifications < 20 ms).\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black hatch marks represent the observed non-local stationary positions. \textbf{(F)} Box plot of the average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}
\caption{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value---which governs the probability of remaining in that state. The corresponding expected duration of staying in the state is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in Figure \ref{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another state. Shaded regions correspond to the classification as in Figure \ref{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental1/Figure2_v3-supplemental1}
\caption{Decoding the same SWR in Figure \ref{2} with 2D position using sorted spikes and clusterless decoding, respectively. \textbf{(A)} The left plot shows the spikes from cells arranged by the linear position of the peak of place field as in Figure 2. The middle plot shows the probability of each dynamic over time from the 2D decode. Shaded regions indicate classification category as in Figure 1G and 2. The rightmost plot shows the most probable estimate of the latent position (MAP estimate) with color indicating time. The latent position posterior summed over time is shown in the purple shading. The light grey lines represent the position of the animal over the entire epoch and the magenta dot represents the animal's position. \textbf{(B)} Same as in (A), but with clusterless decoding.}
\label{fig:Figure2-Figure supplement 1}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2-supplemental2/Figure2_v4-supplemental2}
\caption{A-F. More examples of SWRs that have continuous trajectories. Left panel uses the same conventions as Figure \ref{2}A and \ref{2}B. Right panel shows the 1D MAP estimate projected back to 2D as in Figure \ref{2}C. Color indicates time. Light grey lines indicate the animal's position over the entire epoch. Magenta dot represents the animal's position during the SWR.}
\label{fig:Figure2-Figure supplement 2}
\end{figure*}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental1/Figure3_v4_supplemental1}
\caption{A-I. More examples of SWRs that would not be well-characterized by using the standard decoder. Conventions are the same as in Figure \ref{3}.
}
\label{fig:Figure3-Figure supplement 1}
\end{figure*}

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental2/Figure3_v2-supplemental2}
\caption{Comparison of percentage of SWRs classified (that is, a SWR containing at least one of the five classifications) on real vs. position shuffled data for two epochs from different animals. Red line represents the percent of SWR events classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing and position occupancy.}
\label{fig:Figure3-Figure supplement 2}
\end{figure}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4-supplemental1/Figure4_v1-supplemental1}
\caption{Comparison of average spatial coverage of the 95\% highest posterior density on real vs. position shuffled data for two epochs from different animals. Red line represents the average spatial position spanned by the highest posterior density on real data for all ripples. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing and position occupancy.}
\label{fig:Figure4-Figure supplement 1}
\end{figure*}
